Exploring Local Explanations of Non-linear Models Using Animated Linear Projections

### 2022-01-13 _paper.rmd
- Updates to Di's slack points 1-4, + (6): - Glob view, sq asp ratio - Glob view, reduce jitter - regression facet issue; residual removed.

# Tours and the radial tour

@@

Ames 2018, housing data was subset to North Ames (the neighborhood with the most house sales). The remaining are 338 house sales across nine features. Using interaction from the global view, we select a house with an extreme negative residual and an accurate instance close to it in the data.

XXX You need to explain the variables.

(ref:caseames) Ames housing 2018 regressing sales price [2018 USD]. The PI sale price was under predicted, and had sizable attribution to lot area (`LtA`). The CI was predicted sales price was similar and much more accurate with its observed sales price while it has very little attribution to lot area. Varying the contribution lot area the separation between these house sales crosses when there is a low contribution of `LtA`,  which is important to explaining the PI and near invariant to the sales price of the CI. The corresponding animation is at [shorturl.at/oqyDU](https://github.com/nspyrison/cheem_paper/blob/main/figures/case_ames2018.mp4)
```{r caseames, echo=F, out.width="100%", fig.cap = "(ref:caseames)"}
knitr::include_graphics("./figures/case_ames2018.png")
```

>>>>

# Tours and the radial tour {#sec:tour}

@@

Ames 2018, housing data was subset to North Ames (the neighborhood with the most house sales). The remaining are 338 house sales. A random forest is regressed this price with the features `LotArea` (`LtA`), `OverallQual` (`Qlt`), `YearBuild` (`YrB`), `LivingArea` (`LvA`), `Bathrms` (`Bth`), `Bedrms` (`Bdr`), `TotRms` (`Rms`), `GarageYrBlt` (`GYB`), and `GarageArea` (`GrA`). Using interaction from the global view, we select a house with an extreme negative residual and an accurate instance close to it in the data.

(ref:caseames) Ames housing 2018 regressing sales price [USD]. The PI sale price was under predicted, and had sizable attribution to lot area (`LtA`). The CI was predicted sales price was similar and much more accurate with its observed sales price while it has very little attribution to lot area. Varying the contribution lot area the separation between these house sales crosses when there is a low contribution of `LtA`,  which is important to explaining the PI and near invariant to the sales price of the CI. The corresponding animation is at [shorturl.at/oqyDU](https://github.com/nspyrison/cheem_paper/blob/main/figures/case_ames2018.mp4)
```{r caseames, echo=F, out.width="100%", fig.cap = "(ref:caseames)"}
knitr::include_graphics("./figures/case_ames2018.png")
```

### 2022-01-13 _paper.tex
- Updates to Di's slack points 1-4, + (6): - Glob view, sq asp ratio - Glob view, reduce jitter - regression facet issue; residual removed.

\hypertarget{tours-and-the-radial-tour}{%
\section{Tours and the radial tour}\label{tours-and-the-radial-tour}}

@@

{\centering \includegraphics[width=1\linewidth]{./figures/case_penguins_BlFl}

\caption{An orthogonal view of penguins data showing the PI, a Gentoo (purple) close to the distribution of Chinstrap (orange).}\label{fig:casepenguinsblfl}
\end{figure}

@@

\caption{Examining the SHAP values for a random forest model classifying Palmer penguin species. The PI is an Chinstrap (orange) penguin that is misclassified as a Gentoo (purple), marked as an asterisk in (a), and the dashed vertical line in (b). Starting from the attribution projection the contribution of \texttt{b\_l} is varied. This explanation is not supported sense by the time there is moderate contribution. The basis showing the correct species separation is fairly close to the regularly attribution to other Gentoo penguins aside from the sign change on flipper length. The .mp4 animation of this tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_penguins.mp4}{shorturl.at/hFHNS}}\label{fig:casepenguins}
\end{figure}

Figure \ref{fig:casepenguins} shows plots from the cheem viewer for exploring the random forest model on the penguins data. Plot (a) shows the global view, and plot (b) shows several 1D projections generated with the radial tour. Penguin 243, an Chinstrap (orange), is the PI because it has been misclassified as a Gentoo (purple).

There is more separation visible in the attribution space than the data space, as would be expected. The observed vs predicted plot reveals a handful of misclassified instances. A misclassified point (represented by the asterisk in the global view, and as a dashed vertical line in the tour view) is contrasted with a correctly classified point (represented by an \(\times\) and vertical dotted line).

The radial tour starts from the attribution projection of the misclassified instance (left plot). The important features identifed by SHAP in the (wrong) prediction for this instance is mostly \texttt{b\_l} and \texttt{b\_d} with small contributions of \texttt{f\_l} and \texttt{wgt}. This projection is a view where the Gentoo (purple) looks much more likely for this instance than Chinstrap. For this instance we would like to know why an Adelie has been wrongly labelled a Chinstrap, so this combination of features is not particularly useful. To explore this we use the radial tour to vary the contribution of flipper length (\texttt{f\_l}). The more that \texttt{f\_l} contributes to the the projection the more this penguin looks like a Gentoo, its observed species. This suggests that \texttt{f\_l} should have been an important feature for explaining the (wrong) prediction.

@@

The chocolates dataset consists of 88 instances of ten nutritional measurements determined from their labels, and labeled as being either milk or dark. Dark chocolates are typically considered healthier and often can be marketed at a higher price. It could be interesting to examine the nutritional properties of any dark chocolates that have been misclassified as milk, to determine if a manufacturer is being honest. A random forest model is used, and a chocolate labeled dark but predicted to be milk is chosen as the instance of interest (chocolate 22). It is compared with a correctly classified dark chocolate.

@@

\caption{Chocolates data type classification (milk or dark). We select a chocolate labeled as dark though a random forest model predicts it to be milk chocolate from the values on the nutritional label. The attribution projection already looks more like a dark chocolate than milk. We remove four features with the lowest contribution for the selected instance and vary the contribution of \texttt{sugar}. The misclassification of two instances seems improbable when sugar is near max contribution. Animated tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_chocolates.mp4}{shorturl.at/dBE28}.}\label{fig:casechocolates}
\end{figure}

Figure \ref{fig:casechocolates} similarly shows that attribution-space is more separable than data-space. Interestingly, the class imbalance that we suspected was not observed; there are only six chocolates labeled as dark and predicted as milk, while eight of the inverse case. We look at one extreme instance and vary the contribution of \texttt{sugar}. We see notice two instances that have clearly well within the milk chocolate despite being observed as dark chocolate on their product label. While conjecture, there are plausible reasons that a manufacturer has incentives to cut corners and label their products different than what they are. In the penguin case we explored how the model misclassified an instance from a type one error perspective. In contrast, in the chocolates cases shows a type two error view; looking for extreme support of the models classification to challenge the original instance.

@@

Ames 2018, housing data was subset to North Ames (the neighborhood with the most house sales). The remaining are 338 house sales across nine features. Using interaction from the global view, we select a house with an extreme negative residual and an accurate instance close to it in the data.

@@

\caption{Ames housing 2018 regressing sales price {[}2018 USD{]}. The PI sale price was under predicted, and had sizable attribution to lot area (\texttt{LtA}). The CI was predicted sales price was similar and much more accurate with its observed sales price while it has very little attribution to lot area. Varying the contribution lot area the separation between these house sales crosses when there is a low contribution of \texttt{LtA}, which is important to explaining the PI and near invariant to the sales price of the CI. The corresponding animation is at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_ames2018.mp4}{shorturl.at/oqyDU}}\label{fig:caseames}
\end{figure}

>>>>

\hypertarget{sec:tour}{%
\section{Tours and the radial tour}\label{sec:tour}}

@@

@@

\caption{Examining the SHAP values for a random forest model classifying Palmer penguin species. The PI is an Chinstrap (orange) penguin that is misclassified as a Gentoo (purple), marked as an asterisk in (a), and the dashed vertical line in (b). The radial view shows varying the contribution of \texttt{f\_l} from the initial attribution projection (b, left), which produces a linear combination where the PI is more probably a Chinstrap than a Gentoo (b, right). (The .mp4 animation of the radial tour is at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_penguins.mp4}{shorturl.at/hFHNS}.)}\label{fig:casepenguins}
\end{figure}

Figure \ref{fig:casepenguins} shows plots from the cheem viewer for exploring the random forest model on the penguins data. Plot (a) shows the global view, and plot (b) shows several 1D projections generated with the radial tour. Penguin 243, a Gentoo (purple), is the PI because it has been misclassified as a Chinstrap (orange).

{\centering \includegraphics[width=0.6\linewidth]{./figures/case_penguins_BlFl}

\caption{Checking what is learned from the cheem viewer. This is a plot of flipper length (\texttt{f\_l}) and bill length (\texttt{b\_l}), where the PI is highlighted by an asterisk. It is a Gentoo (purple) misclassified as a Chinstrap (orange). The PI has an unusually small \texttt{f\_l} length which is the primary reason it is confused with a Chinstrap.}\label{fig:casepenguinsblfl}
\end{figure}

There is more separation visible in the attribution space than the data space, as would be expected. The observed vs predicted plot reveals a handful of misclassified instances. For this illustration we are going to explore why a Gentoo has been wrongly labelled as a Chinstrap. The PI is a misclassified point (represented by the asterisk in the global view, and as a dashed vertical line in the tour view). The CI is a correctly classified point (represented by an \(\times\) and a vertical dotted line).

The radial tour starts from the attribution projection of the misclassified instance (b, left). The important features identifed by SHAP in the (wrong) prediction for this instance is mostly \texttt{b\_l} and \texttt{b\_d} with small contributions of \texttt{f\_l} and \texttt{wgt}. This projection is a view where the Gentoo (purple) looks much more likely for this instance than Chinstrap, that is, this combination of features is not particularly useful because the PI looks very much like other Gentoo penguins. To explore this we use the radial tour to vary the contribution of flipper length (\texttt{f\_l}). (In our exploreation, this was the third variable explored. It is typically useful to explore the variables with larger contributions, here \texttt{b\_l} and \texttt{b\_d}, but when doing this nothing was revealed about how the PI differed from other Gentoos.) On varying \texttt{f\_l} as it contributes more to the projection (b, right) we see that more, and more, this penguin looks like a Chinstrap. This suggests that \texttt{f\_l} should be considered to be an important feature for explaining the (wrong) prediction.

Figure \ref{fig:casepenguinsblfl} confirms that flipper length (\texttt{f\_l}) is important for the confusion of the PI as a Chinstrap. Here, flipper length and body length are plotted and we can see that the PI is closer to the Chinstrap group in these two variables, mostly because it has an unusually low value of flipper length relative to other Gentoos.

@@

The chocolates dataset consists of 88 observations of ten nutritional measurements determined from their labels, and labeled as being either milk or dark. Dark chocolate is considered healthier than milk. The data was collected by students during the Iowa State University class STAT503 from nutritional information from the manufacturer's web site, and normalized to 100g equivalents. The data is available in the \texttt{cheem} package. A random forest model is used for the classification of chocolate type.

It could be interesting to examine the nutritional properties of any dark chocolates that have been misclassified as milk. A reason to do this is that a dark chocolate that is nutritionally more like milk should not be considered as a healthy alternative. It is interesting to explore which of the nutritional variables contribute most to the misclassification.

@@

\caption{Chocolates data type classification (milk or dark). We select a chocolate labeled as dark though a random forest model predicts it to be milk chocolate from the values on the nutritional label. The attribution projection already looks more like a dark chocolate than milk. We remove a 4 variables with the lowest contribution for the selected instance and vary the contribution of \texttt{fiber}. The misclassification of two instances seems improbable when sugar is near max contribution. Animated tour can be found at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_chocolates.mp4}{shorturl.at/dBE28}.}\label{fig:casechocolates}
\end{figure}

This type of exploration is shown in \ref{fig:casechocolates}, where a chocolate labeled dark but predicted to be milk is chosen as the observation of interest (observation 22). It is compared with a correctly classified dark chocolate (observation 7). The PCA plot, and the SHAP PCA plots show a big difference between the two chocolate types but with confusion for a handful of observations. The misclassifications are clearer in the observed vs predicted plot, and can be seens to be mistaken in both ways: milk to dark and dark to milk.

The attribution projection for chocolate 22 suggests that Fiber, Sugars and Calories are most responsible for its incorrect prediction. The way to read this plot is to see that Fiber has a large negative value, while Sugars and Calories have reasonably large positive values. In the density plot, observations on the very left of the display would have high values of Fiber (matching the negative projection coefficient) and low values of Sugars and Calories. The opposite would be the interpretation of a point with high values in this plot. The dark chocolates (orange) are mostly on the left, and this is a reason why they are considered to be healthier: high fibre and low sugar. The density for milk chocolates is further to the right, indicating that they generally have low fibre and high sugar.

The observation of interest (dashed line) can be viewed against the comparison observation (dotted line). Now one needs to pay different attention to the parallel plot of the SHAP values, which are local to a particular observation, and the density plot which is the same projection of all observations as specifed by the SHAP values of the observation of interest.

From the parallel coordinate plot we can quickly compare the variable contributions to the two different predictions. The observation of interest differs form the comparison primarily on the Fiber variable, which suggests that this is the reason for the incorrect prediction.

From the density plot, which is the attribution projection corresponding to the observation of interest, both observations are more like dark chocolates. If we vary the contribution of Sugars, and completely remove Sugars from the projection, this is where the difference becomes apparent. When primarily Fiber is examined observation 22 looks more like a milk chocolate.

@@

Ames 2018, housing data was subset to North Ames (the neighborhood with the most house sales). The remaining are 338 house sales. A random forest is regressed this price with the features \texttt{LotArea} (\texttt{LtA}), \texttt{OverallQual} (\texttt{Qlt}), \texttt{YearBuild} (\texttt{YrB}), \texttt{LivingArea} (\texttt{LvA}), \texttt{Bathrms} (\texttt{Bth}), \texttt{Bedrms} (\texttt{Bdr}), \texttt{TotRms} (\texttt{Rms}), \texttt{GarageYrBlt} (\texttt{GYB}), and \texttt{GarageArea} (\texttt{GrA}). Using interaction from the global view, we select a house with an extreme negative residual and an accurate instance close to it in the data.

@@

\caption{Ames housing 2018 regressing sales price {[}USD{]}. The PI sale price was under predicted, and had sizable attribution to lot area (\texttt{LtA}). The CI was predicted sales price was similar and much more accurate with its observed sales price while it has very little attribution to lot area. Varying the contribution lot area the separation between these house sales crosses when there is a low contribution of \texttt{LtA}, which is important to explaining the PI and near invariant to the sales price of the CI. The corresponding animation is at \href{https://github.com/nspyrison/cheem_paper/blob/main/figures/case_ames2018.mp4}{shorturl.at/oqyDU}}\label{fig:caseames}
\end{figure}
